---
title: "R Session 1 Homework"
author: "Marc Elosua-Bayés"
date: "2/24/2022"
output: html_document
---

Hello and welcome to your first R homework assignment! You should reference both
your notes from the session and the [Base-R cheatsheet](https://github.com/rstudio/cheatsheets/blob/main/base-r.pdf) 
to answer the following questions.

## Part 1: Objects + Basic Math
*Remember to reference the [cheat sheet](https://github.com/rstudio/cheatsheets/blob/main/base-r.pdf)
 if you get stuck*  
 
1. Generate a sequence from 3 to 71 and name it `my_sequence.`  
```{r}

```

2. Generate a sequence from 3 to 71 that only has odd numbers, and call it `odds`.  
```{r}

```

3. Find the sum of `odds`  
```{r}

```

4. Find the product of the 7th and 11th element of `odds`  
```{r}

```

## Part 2: `for` Loops and `if` statements
*Remember to reference the [cheat sheet](https://github.com/rstudio/cheatsheets/blob/main/base-r.pdf)
if you get stuck*  
 
5. Write a for loop that prints out every fifth element of `odds` (i.e., the 
5th element, the 10th element, the 15th element, etc)
```{r}

```

6. Using a for loop, construct a vector of length 10, where the first element 
is `100`, and every following element is double the square-root of the previous 
one.
```{r}
n <- 100
vec <- c()

for (i in 1:10) {
    
    if (i == 1) {
        vec <- c(vec, n)
    } else {
        vec <- c(vec, 2 * sqrt(vec[i - 1]))
    }
   
}
vec
```

7. Write an `if / else` statement that prints "Yay" if some value `x` is 
greater than 10 and "Awesome" otherwise. Try setting `x` to 
different values and try running it.
```{r}
x <- 5

if () {
    
}
```

## Part 3: Functions
*Remember to reference the [cheat sheet](https://github.com/rstudio/cheatsheets/blob/main/base-r.pdf)
 if you get stuck*  
 
Functions are defined like:
```r
name_of_function <- function(parameter) {
  # do_something...
}
```

Here's an example:
```r
foo <- function(x) {
  x * 2 + 3 / 4
}
```

The last calculation that happens in the function will be returned by default. 
If you want, you can add an explicit `return` statement, like this:

```r
foo <- function(x) {
  return(x * 2 + 3 / 4)
}
```
or
```r
foo <- function(x) {
  y <- x * 2 + 3 / 4
  return(y)
}
```
But you don't need to - all three of those examples will work exactly the same.

8. Try it out! Define the above `foo` function in your environment, and try 
running it with different parameters.
```{r}

```

9. Add a second parameter (`y`) to `foo` that raises `x` to the `yth` power.
```{r}

```

10. Moving on from `foo`, define a new function `foo2`, that returns `TRUE` if 
the input is greater than 5.
```{r}

```

To set a default value in a function, define it like:
```r
foo <- function(x = 1) {
  x * 2 + 3 / 4
}
```

11. Define and run the above `foo` function in your environment, and try running
it without any parameters `foo()` and with other parameters `foo(3)`.
```{r}

```

12. Edit your `foo2` function definition to have a default parameter value of 3.
```{r}

```
 
## Part 4: Data
*Remember to reference the [cheat sheet](https://rstudio.com/wp-content/uploads/2016/10/r-cheat-sheet-3.pdf)
 if you get stuck*  

Run the following code in your R console:
```r
library(bigrquery)
library(tidyverse)
library(nytslider)

# Here we're reading in data from BQ
df <- read_bq(
  "SELECT dashboard_desk, EXTRACT(MONTH FROM _pt) month, sum(page_views) pvs
  FROM nyt-bigquery-beta-workspace.audience_data.behavior_counts_view
  WHERE EXTRACT(YEAR FROM _pt) = 2020 AND dashboard_desk IS NOT NULL
  GROUP BY 1, 2")
```

13. Print out the first 6 rows of `df`

14. How many unique `dashboard_desk` values are in df?

15. What's the median `pvs` value where the desk is "Washington"?

16. How many pageviews did "Games" have in the second half of the year?

You can add a new column to a dataframe using the `$`. Try this example:
```r
df$extra <- df$pvs * 2
```

17. Make a new column in `df` called `pvs_halved` that's the `pvs` column 
divided by 2.

18. Make a new column in `df` that's the product of the `pvs` column and the 
`month` column.  

## Part 5: Putting it all together

19. Write a function that takes in (1) a dataframe and (2) a desk name as inputs, 
and returns the original dataframe filtered to that desk - in other words, a 
function that does the r equivalent to:
```SQL
SELECT * FROM {original_table} WHERE dashboard_desk = '{desk}'
```

20. Write a function that takes in (1) a dataframe, (2) a desk name, and (3) a 
month, that returns the sum of pageviews for that desk, after that month - in 
other words, a function that does the r equivalnet to:
```SQL
SELECT SUM(pvs) FROM {original_table} 
WHERE dashboard_deks = '{desk}' AND month >= {month_value}
```

21. Using a `for` loop, print out the name of every `dashboard_desk` in `df`. 
(*hint: use the `unique()` function*)  

22. Using the `sort()` function, modify your for loop to print out the name of 
every `dashboard_desk` in `df` in alphabetical order.

23. Now modify your for loop even further: iterate through the desks (in 
alphabetical order) and for each desk, print out the name of the desk and the 
total sum of that desk's pageviews in `df`.

## THAT'S ALL!

I'll send around an answer key in a few days, but take a stab at getting through
 what you can on your own. And if you have time, read Hadley Wickham's paper 
 on [tidy data](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html), 
 and then you will be all set for Session 2!